---
title: "STA 207 Project Star"
author: "Danny Kuei"
date: "2026-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(haven)
library(ggplot2)
library(dplyr)
```

```{r}
df_highschools <- read_sav("Documents/STA 207/Project/dataverse_files/PROJECT STAR/STAR_High_Schools.sav")
df_comp_students <- read_sav("Documents/STA 207/Project/dataverse_files/PROJECT STAR/Comparison_Students.sav")
df_students <- read_sav("Documents/STA 207/Project/dataverse_files/PROJECT STAR/STAR_Students.sav")
df_k3schools <- read_sav("Documents/STA 207/Project/dataverse_files/PROJECT STAR/STAR_K-3_Schools.sav")
```

```{r}
table(df_students$gkclasstype)
```

# Data Wrangling

## Dataframe filters to check for sample size
```{r}
# Filter for no aide
had_aide <- df_students$gkclasstype == 3 | 
  df_students$g1classtype == 3 | 
  df_students$g2classtype == 3 | 
  df_students$g3classtype == 3

df_students_noaide <- df_students[!had_aide,]
nrow(df_students_noaide)
```

```{r}
# Filter for no aide and were in study all 4 years
df_students_noaide_4years <- df_students[!had_aide &
                                         !is.na(df_students$gkclasstype) & 
                                         !is.na(df_students$g1classtype) & 
                                         !is.na(df_students$g2classtype) & 
                                         !is.na(df_students$g3classtype), ]

nrow(df_students_noaide_4years)
```



## Getting class type groups
```{r}
df_students_noaide_4years$class_group <- "Other" # initialize new columns

# move column
df_students_noaide_4years <- df_students_noaide_4years %>% 
  relocate(class_group, .before = gkclasstype) %>%
  relocate(yearssmall, .before = gkclasstype)
```

```{r}
table(df_students$yearssmall)
table(df_students_noaide_4years$yearssmall)
```

```{r}
table(df_students$gkclasstype)
table(df_students$g1classtype)
table(df_students$g2classtype)
table(df_students$g3classtype)

table(df_students_noaide_4years$gkclasstype)
table(df_students_noaide_4years$g1classtype)
table(df_students_noaide_4years$g2classtype)
table(df_students_noaide_4years$g3classtype)

```


```{r}
# Change class type groups for each student

# small 4 years
ssss <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[ssss] <- "ssss"

# small first 3 years, regular last year
sssr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[sssr] <- "sssr"


# small first 2 years, regular last 2 years
ssrr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[ssrr] <- "ssrr"

# small first year, regular last 3 years
srrr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[srrr] <- "srrr"

# regular 4 years
rrrr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rrrr] <- "rrrr"

# regular first 3 years, small last year
rrrs <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rrrs] <- "rrrs"

# regular first 2 year, small last 2 years
rrss <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rrss] <- "rrss"

# regular first year, small last 3 years
rsss <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rsss] <- "rsss"

# small 2 years, regular 1 year, then small last year
ssrs <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[ssrs] <- "ssrs"

# small first 1 year, regular 2 years, then small last year
srrs <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[srrs] <- "srrs"


# small first year, regular, small, regular
srsr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[srsr] <- "srsr"

# small first year, regular, small, small
srss <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[srss] <- "srss"

# regular first 2 years, small, regular
rrsr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rrsr] <- "rrsr"

# regular first year, small, small, regular
rssr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rssr] <- "rssr"

# regular first year, small, regular, small
rsrs <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rsrs] <- "rsrs"

# regular first year, small, regular, regular
rsrr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rsrr] <- "rsrr"


```


```{r}
table(df_students_noaide_4years$class_group)
```
The subset of all small and all regular has good sample size. The subsets of regular to small (rrrs, rrss, rsss) are also sufficient in which we could compare if multiple years leads to higher scores and we could comapre 3rd grade. However subsetting For the other way around small to regular (srrr, ssrr, sssr), the sample is lower. 


## Class Type "Trend" Groups
```{r}
# initalize trend columns
df_students_noaide_4years$trend_all <- "Other"
df_students_noaide_4years$trend_k_to_1 <- "Other"
df_students_noaide_4years$trend_1_to_2 <- "Other"
df_students_noaide_4years$trend_2_to_3 <- "Other"

# move column
df_students_noaide_4years <- df_students_noaide_4years %>% 
  relocate(trend_all, .before = class_group) %>%
  relocate(trend_k_to_1, .before = class_group) %>%
  relocate(trend_1_to_2, .before = class_group) %>%
  relocate(trend_2_to_3, .before = class_group)
```

```{r}
# Change class trend groups for each student

# All 4 years of small vs regular has enough sample size so we can keep those the same
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group == "ssss"] <- "constant_small"
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group == "rrrr"] <- "constant_reg"

# Small to regular trend (srrr, ssrr, sssr)
subset <- c('srrr', 'ssrr', 'sssr')
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group %in% subset] <- "small_to_reg"

# Regular to small trend (rsss, rrss, rrrs)
subset <- c('rsss', 'rrss', 'rrrs')
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group %in% subset] <- "reg_to_small"
```


# Cumulative Effects Results
## Compare Grade 3 scores approach
```{r}
# Subset for groups of interest (Regular --> Small)
subset <- c('ssss', 'rsss', 'rrss', 'rrrs', 'rrrr')

df <- df_students_noaide_4years[df_students_noaide_4years$class_group %in% subset, ]


# Plot by trend (Regular --> Small)

# math
ggplot(df, aes(x = class_group, y = g3tmathss)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  labs(title = "Class Size Trend on Grade 3 Math",
       x = "Sequence (R=Regular, S=Small)", y = "Mean Scale SAT Score")

# reading
ggplot(df, aes(x = class_group, y = g3treadss)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  labs(title = "Class Size Trend on Grade 3 Reading",
       x = "Sequence (R=Regular, S=Small)", y = "Mean Scaled SAT Score")

# table
aggregate(cbind(g3tmathss, g3treadss, yearssmall) ~ class_group, 
                           data = df, 
                           FUN = mean, 
                           na.rm = TRUE)
```



```{r}
# Subset for groups of interest (Small --> Regular)
subset <- c('srrr', 'ssrr', 'sssr', 'rrrr')

df <- df_students_noaide_4years[df_students_noaide_4years$class_group %in% subset, ]


# Plot by trend (Regular --> Small)

# math
ggplot(df, aes(x = class_group, y = g3tmathss)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  labs(title = "Class Size Trend on Grade 3 Math",
       x = "Sequence (R=Regular, S=Small)", y = "Mean Scale SAT Score")

# reading
ggplot(df, aes(x = class_group, y = g3treadss)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  labs(title = "Class Size Trend on Grade 3 Reading",
       x = "Sequence (R=Regular, S=Small)", y = "Mean Scaled SAT Score")

# table
aggregate(cbind(g3tmathss, g3treadss, yearssmall) ~ class_group, 
                           data = df, 
                           FUN = mean, 
                           na.rm = TRUE)
```



```{r}
# Subset for groups of interest (All)
subset <- c('ssss', 'rsss', 'rrss', 'rrrs', 'rrrr', 'srrr', 'ssrr', 'sssr', 'srss', 'srrs', 'rssr', 'rsrr')

df <- df_students_noaide_4years[df_students_noaide_4years$class_group %in% subset, ]


# Plot by cumulative years

# math
ggplot(df, aes(x = yearssmall, y = g3tmathss)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  labs(title = "Number of Year of Small Class vs Grade 3 Math",
       x = "Sequence (R=Regular, S=Small)", y = "Mean Scale SAT Score")


# reading
ggplot(df, aes(x = yearssmall, y = g3treadss)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  labs(title = "Number of Year of Small Class vs Grade 3 Reading",
       x = "Sequence (R=Regular, S=Small)", y = "Mean Scaled SAT Score")

# table
aggregate(cbind(g3tmathss, g3treadss) ~ yearssmall, 
                           data = df, 
                           FUN = mean, 
                           na.rm = TRUE)

```

```{r}
# By trend

# math
ggplot(df_students_noaide_4years, aes(x = trend_all, y = g3tmathss)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  labs(title = "Class Size Trend on Grade 3 Math",
       x = "Sequence (R=Regular, S=Small)", y = "Mean Scale SAT Score")

# reading
ggplot(df_students_noaide_4years, aes(x = trend_all, y = g3treadss)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  labs(title = "Class Size Trend on Grade 3 Reading",
       x = "Sequence (R=Regular, S=Small)", y = "Mean Scaled SAT Score")

# table
aggregate(cbind(g3tmathss, g3treadss) ~ trend_all, 
                           data = df_students_noaide_4years, 
                           FUN = mean, 
                           na.rm = TRUE)
```



## Weighted Average Approach
```{r}
# For now, just take mean scores
df_students_noaide_4years$math_mean <- rowMeans(df_students_noaide_4years[, c("gktmathss", "g1tmathss", "g2tmathss", "g3tmathss")], na.rm = TRUE)
df_students_noaide_4years$read_mean <- rowMeans(df_students_noaide_4years[, c("gktreadss", "g1treadss", "g2treadss", "g3treadss")], na.rm = TRUE)

grade_means <- aggregate(cbind(gktmathss, g1tmathss, g2tmathss, g3tmathss) ~ yearssmall, 
                         data = df_students_noaide_4years, 
                         FUN = mean, na.rm = TRUE)

df_long <- reshape(grade_means, 
                   varying = c("gktmathss", "g1tmathss", "g2tmathss", "g3tmathss"), 
                   v.names = "Mean_Score", 
                   timevar = "Grade", 
                   times = c("K", "1", "2", "3"), 
                   direction = "long")

# math
ggplot(df_students_noaide_4years, aes(x = yearssmall, y = math_mean)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  labs(title = "Class Size Years on Mean Grades",
       x = "Sequence (R=Regular, S=Small)", y = "Mean Scale SAT Score")

# reading
ggplot(df_students_noaide_4years, aes(x = yearssmall, y = read_mean)) +
  stat_summary(fun = mean, geom = "bar") +
  theme_minimal() +
  labs(title = "Class Size Years on Mean Grades",
       x = "Sequence (R=Regular, S=Small)", y = "Mean Scaled SAT Score")

# table
aggregate(cbind(math_mean, read_mean) ~ trend_all, 
                           data = df_students_noaide_4years, 
                           FUN = mean, 
                           na.rm = TRUE)

```



# Looking at other factors (Socioeconomic, Gender, Race)

## Number of years small class
```{r}
# Set factors and labels
df_students_noaide_4years$gender <- factor(df_students_noaide_4years$gender, labels = c("Male", "Female")) # gender

df_students_noaide_4years <- subset(df_students_noaide_4years, race %in% c(1,2))
df_students_noaide_4years$race <- factor(df_students_noaide_4years$race, labels = c("White", "Black")) # ethnicity

df_students_noaide_4years$gkfreelunch <- factor(df_students_noaide_4years$gkfreelunch, labels = c("Free", "Non-Free")) # free lunch
```

```{r}
# additive model with socioeconmic (lunch free), years small, ethnicity, gender
model_add <- lm(g3tmathss ~ yearssmall + gender + race + gkfreelunch, data = df_students_noaide_4years)
summary(model_add)
```

```{r}
# interactions model with socioeconmic (lunch free), years small, ethnicity, gender, yearssmall * race
model_int1 <- lm(g3tmathss ~ yearssmall*race + gender + gkfreelunch, data = df_students_noaide_4years)
summary(model_int1)
```

```{r}
# interactions model with socioeconmic (lunch free), years small, ethnicity, gender, years small * gkfreelunch
model_int2 <- lm(g3tmathss ~ yearssmall*gkfreelunch + gender + race, data = df_students_noaide_4years)
summary(model_int2)
```

```{r}
# interactions model with socioeconmic (lunch free), years small, ethnicity, gender, years small * gkfreelunch
model_int3 <- lm(g3tmathss ~ yearssmall*gkfreelunch + yearssmall*race + gender, data = df_students_noaide_4years)
summary(model_int3)
```

```{r}
# Try AIC for best model
AIC(model_add, model_int1)
AIC(model_add, model_int2)
AIC(model_add, model_int3)
```

```{r}
# Try ANOVA for best model
anova(model_add, model_int1)
anova(model_add, model_int2)
anova(model_add, model_int3)
```

By both AIC and likelihood ratio tests, it seems additive model is sufficient.

## Small class pattern
```{r}
# if we are examining grade 3 math, we have to look at students of similar groups
# Subset for groups of interest (Regular --> Small)
subset <- c('ssss', 'rsss', 'rrss', 'rrrs', 'rrrr')

df <- df_students_noaide_4years[df_students_noaide_4years$class_group %in% subset, ]
nrow(df)
```

```{r}
# additive model with socioeconmic (lunch free), years small, ethnicity, gender
model_add <- lm(g3tmathss ~ class_group + gender + race + gkfreelunch, data = df)
summary(model_add)
```

```{r}
# interactions model with socioeconmic (lunch free), years small, ethnicity, gender, yearssmall * race
model_int1 <- lm(g3tmathss ~ class_group*race + gender + gkfreelunch, data = df)
summary(model_int1)
```

```{r}
# interactions model with socioeconmic (lunch free), years small, ethnicity, gender, years small * gkfreelunch
model_int2 <- lm(g3tmathss ~ class_group*gkfreelunch + gender + race, data = df)
summary(model_int2)
```

```{r}
# interactions model with socioeconmic (lunch free), years small, ethnicity, gender, years small * gkfreelunch
model_int3 <- lm(g3tmathss ~ class_group*gkfreelunch + class_group*race + gender, data = df)
summary(model_int3)
```

```{r}
# Try AIC for best model
AIC(model_add, model_int1)
AIC(model_add, model_int2)
AIC(model_add, model_int3)
```

```{r}
# Try ANOVA for best model
anova(model_add, model_int1)
anova(model_add, model_int2)
anova(model_add, model_int3)
```

  



