---
title: "STA 207 Project Star"
author: "Adam Hetherwick & Danny Kuei"
date: "2026-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(haven)
library(ggplot2)
library(dplyr)
```

```{r, include=FALSE}
df_highschools <- read_sav("dataverse_files/STAR_High_Schools.sav")
df_comp_students <- read_sav("dataverse_files/Comparison_Students.sav")
df_students <- read_sav("dataverse_files/STAR_Students.sav")
df_k3schools <- read_sav("dataverse_files/STAR_K-3_Schools.sav")
```

```{r}
table(df_students$gkclasstype)
```

# Data Wrangling

## Dataframe filters to check for sample size
```{r}
# Filter for no aide
had_aide <- df_students$gkclasstype == 3 | 
  df_students$g1classtype == 3 | 
  df_students$g2classtype == 3 | 
  df_students$g3classtype == 3

df_students_noaide <- df_students[!had_aide,]
nrow(df_students_noaide)
```

```{r}
# Filter for no aide and were in study all 4 years
df_students_noaide_4years <- df_students[!had_aide &
                                         !is.na(df_students$gkclasstype) & 
                                         !is.na(df_students$g1classtype) & 
                                         !is.na(df_students$g2classtype) & 
                                         !is.na(df_students$g3classtype), ]

nrow(df_students_noaide_4years)
```

## Getting class type groups (scrapped due to sample size)

```{r}
df_students_noaide_4years$class_group <- "Other" # initialize new columns

# move column
df_students_noaide_4years <- df_students_noaide_4years %>% 
  relocate(class_group, .before = gkclasstype) %>%
  relocate(yearssmall, .before = gkclasstype)
```

```{r}
# Change class type groups for each student

# small 4 years
ssss <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[ssss] <- "ssss"

# small first 3 years, regular last year
sssr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[sssr] <- "sssr"


# small first 2 years, regular last 2 years
ssrr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[ssrr] <- "ssrr"

# small first year, regular last 3 years
srrr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[srrr] <- "srrr"

# regular 4 years
rrrr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rrrr] <- "rrrr"

# regular first 3 years, small last year
rrrs <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rrrs] <- "rrrs"

# regular first 2 year, small last 2 years
rrss <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rrss] <- "rrss"

# regular first year, small last 3 years
rsss <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rsss] <- "rsss"

# small 2 years, regular 1 year, then small last year
ssrs <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[ssrs] <- "ssrs"

# small first 1 year, regular 2 years, then small last year
srrs <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[srrs] <- "srrs"


# small first year, regular, small, regular
srsr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[srsr] <- "srsr"

# small first year, regular, small, small
srss <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[srss] <- "srss"

# regular first 2 years, small, regular
rrsr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rrsr] <- "rrsr"

# regular first year, small, small, regular
rssr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rssr] <- "rssr"

# regular first year, small, regular, small
rsrs <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rsrs] <- "rsrs"

# regular first year, small, regular, regular
rsrr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rsrr] <- "rsrr"
```

## Class Type "Trend" Groups

```{r}
# initalize trend column
df_students_noaide_4years$trend_all <- "Other"

# move column
df_students_noaide_4years <- df_students_noaide_4years %>% 
  relocate(trend_all, .before = class_group) 
```

```{r}
# Change class trend groups for each student

# All 4 years of small vs regular has enough sample size so we can keep those the same
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group == "ssss"] <- "constant_small"
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group == "rrrr"] <- "constant_reg"

# Small to regular trend (srrr, ssrr, sssr)
subset <- c('srrr', 'ssrr', 'sssr')
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group %in% subset] <- "small_to_reg"

# Regular to small trend (rsss, rrss, rrrs)
subset <- c('rsss', 'rrss', 'rrrs')
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group %in% subset] <- "reg_to_small"
```

## Factor Labels

```{r}
# Set factors and labels

# Urban and free lunch variables for question 1 
df_students_noaide_4years$gksurban <- factor(df_students_noaide_4years$gksurban, labels = c("Inner City", "Suburban", "Rural", "Urban")) # urban


# Gender, race, and free lunch variables for question 2
df_students_noaide_4years$gender <- factor(df_students_noaide_4years$gender, labels = c("Male", "Female")) # gender

df_students_noaide_4years <- subset(df_students_noaide_4years, race %in% c(1,2)) # subset for white and black only
df_students_noaide_4years$race <- factor(df_students_noaide_4years$race, labels = c("White", "Black")) # race

df_students_noaide_4years$g3freelunch <- factor(df_students_noaide_4years$g3freelunch, labels = c("Free", "Non-Free")) # free lunch for grade 3
df_students_noaide_4years$gkfreelunch <- factor(df_students_noaide_4years$gkfreelunch, labels = c("Free", "Non-Free")) # free lunch for kindergarten
```

```{r}
nrow(df_students_noaide_4years)
```

# Early Year Class Size vs Socioeconomic (Question 1)

## Three Way Anova Model & F Test

```{r}
library(knitr)
library(kableExtra)

# Urbanicity + free lunch
model_fixed1 <- aov(gktmathss ~ gkclasstype + gkfreelunch + gksurban, data = df_students_noaide_4years)

anova_table = as.data.frame(summary(model_fixed1)[[1]])
anova_table$F_critical = c(3.85, 3.85, 2.61, NA)
anova_table <- anova_table[, c("Df", "Sum Sq", "Mean Sq", "F value", "F_critical", "Pr(>F)")]
anova_table$`Pr(>F)` = c("1.41e-05 ***", "4.47e-12 ***", "0.74 .", "NA")
rownames(anova_table) = c("Class Type", "Free Lunch", "School Location", "Residuals")
kable(anova_table, 
      caption = "Table 4: ANOVA Table for class size, free lunch, and urbanicity", 
      digits=3,
      align = c("l", "l", "l", "l")) %>%
  kable_classic(html_font = "Cambria")

paste("Class Size F_critical: ", round(qf(0.95, 1, 1336), 2))
paste("Free Lunch F_critical: ", round(qf(0.95, 1, 1336), 2))
paste("Urbanicity F_critical: ", round(qf(0.95, 3, 1336), 2))
```

```{r}
# Plot math scores grouped by class type, free lunch, urbanicity?

boxplot(gktmathss ~ gkclasstype,
        data = df_students_noaide_4years,
        main = "Kindergarten Math Scores by Class Type",
        xlab = "Class Type",
        ylab = "Math Score")

boxplot(gktmathss ~ gkfreelunch,
        data = df_students_noaide_4years,
        main = "Kindergarten Math Scores by Free Lunch",
        xlab = "Free Lunch",
        ylab = "Math Score")

boxplot(gktmathss ~ gksurban,
        data = df_students_noaide_4years,
        main = "Kindergarten Math Scores by School Location",
        xlab = "School Location",
        ylab = "Math Score")
```

# Looking at Switch (Question 2)

## Small class pattern

```{r}
# if we are examining grade 3 math, we have to look at students of similar groups
# Subset for groups of interest: Compliers (constant class size) and Noncompliers (switch)
subset <- c('rsss', 'rrss', 'rrrs', 'rrrr', 'ssss', 'srrr', 'ssrr', 'sssr')
df <- df_students_noaide_4years[df_students_noaide_4years$class_group %in% subset, ] 

nrow(df)
```

## Model Selection

```{r}
library(lme4)
library(lmerTest)

# Baselines are constant regular class size, white, male, non-free lunch and teacher experience included

# Mixed effects model where school is random intercept examining trend variable
model_mixed1 <- lmer(g3tmathss ~ trend_all + gender + race + g3freelunch + g3tyears + (1 | g3schid), data = df) # additive
model_mixed2 <- lmer(g3tmathss ~ trend_all * g3freelunch + gender + race + g3tyears + (1 | g3schid), data = df) # interaction trend * lunch
model_mixed3 <- lmer(g3tmathss ~ trend_all * race + gender + g3freelunch + g3tyears + (1 | g3schid), data = df) # interaction trend * race
```

```{r}
# Looking at larger dataset so use BIC for parsomonious model
BIC(model_mixed1, model_mixed2, model_mixed3)
```

```{r}
# Likelihood ratio test
anova(model_mixed1, model_mixed2)
anova(model_mixed1, model_mixed3)
```

Although the interactions model with lunch had a slightly lower BIC, the likelihood ratio test was not significant indicating the additive model is sufficient.

```{r}
summary(model_mixed1)
```

```{r}
anova(model_mixed1)
confint(model_mixed1)
```

## Model Diagnostics

```{r}
plot(model_mixed1,
     main = "Residuals vs. Fitted",
     ylab = "Pearson Residuals",
     xlab = "Fitted Values")
```

The residuals vs fitted plot shows a large clump around 0 with seemingly equal variance throughout, however we notice diagonal lines in the upper right part of the plot. This is due to students having the same combination of gender, race, free lunch, trend, and teacher experience. It is not an issue because we still have no strong pattern in plot, this occurs when we have discrete fitted values. 

```{r}
hist(resid(model_mixed1),
     breaks=25,
     probability = TRUE,
     main = "Residual Distribution",
     xlab = "Residuals")

x <- seq(min(resid(model_mixed1)),
         max(resid(model_mixed1)),
         length = 100)

y <- dnorm(x,
           mean = mean(resid(model_mixed1)),
           sd = sd(resid(model_mixed1)))

lines(x, y, col = "red", lwd = 2, lty = 2)
```

```{r}
plot(scale(resid(model_mixed1)),
     ylab = "Standardized Residuals",
     xlab = "Student Index",
     main = "Studentized Residual Plot")
abline(h=c(-3,3), lty=2)
```


