---
title: "STA 207 Results"
author: "Danny Kuei"
date: "2026-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(haven)
library(ggplot2)
library(dplyr)
library(kableExtra)
```

```{r}
df_highschools <- read_sav("Documents/STA 207/Project/dataverse_files/PROJECT STAR/STAR_High_Schools.sav")
df_comp_students <- read_sav("Documents/STA 207/Project/dataverse_files/PROJECT STAR/Comparison_Students.sav")
df_students <- read_sav("Documents/STA 207/Project/dataverse_files/PROJECT STAR/STAR_Students.sav")
df_k3schools <- read_sav("Documents/STA 207/Project/dataverse_files/PROJECT STAR/STAR_K-3_Schools.sav")
```

```{r}
table(df_students$gkclasstype)
```

# Data Wrangling

## Dataframe filters to check for sample size
```{r}
# Filter for no aide
had_aide <- df_students$gkclasstype == 3 | 
  df_students$g1classtype == 3 | 
  df_students$g2classtype == 3 | 
  df_students$g3classtype == 3

df_students_noaide <- df_students[!had_aide,]
nrow(df_students_noaide)
```

```{r}
# Filter for no aide and were in study all 4 years
df_students_noaide_4years <- df_students[!had_aide &
                                         !is.na(df_students$gkclasstype) & 
                                         !is.na(df_students$g1classtype) & 
                                         !is.na(df_students$g2classtype) & 
                                         !is.na(df_students$g3classtype), ]

nrow(df_students_noaide_4years)
```



## Getting class type groups (scrapped due to sample size)
```{r}
df_students_noaide_4years$class_group <- "Other" # initialize new columns

# move column
df_students_noaide_4years <- df_students_noaide_4years %>% 
  relocate(class_group, .before = gkclasstype) %>%
  relocate(yearssmall, .before = gkclasstype)
```

```{r}
table(df_students$yearssmall)
table(df_students_noaide_4years$yearssmall)
```

```{r}
table(df_students$gkclasstype)
table(df_students$g1classtype)
table(df_students$g2classtype)
table(df_students$g3classtype)

table(df_students_noaide_4years$gkclasstype)
table(df_students_noaide_4years$g1classtype)
table(df_students_noaide_4years$g2classtype)
table(df_students_noaide_4years$g3classtype)
```


```{r}
# Change class type groups for each student

# small 4 years
ssss <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[ssss] <- "ssss"

# small first 3 years, regular last year
sssr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[sssr] <- "sssr"


# small first 2 years, regular last 2 years
ssrr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[ssrr] <- "ssrr"

# small first year, regular last 3 years
srrr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[srrr] <- "srrr"

# regular 4 years
rrrr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rrrr] <- "rrrr"

# regular first 3 years, small last year
rrrs <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rrrs] <- "rrrs"

# regular first 2 year, small last 2 years
rrss <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rrss] <- "rrss"

# regular first year, small last 3 years
rsss <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rsss] <- "rsss"

# small 2 years, regular 1 year, then small last year
ssrs <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[ssrs] <- "ssrs"

# small first 1 year, regular 2 years, then small last year
srrs <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[srrs] <- "srrs"


# small first year, regular, small, regular
srsr <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[srsr] <- "srsr"

# small first year, regular, small, small
srss <-  df_students_noaide_4years$gkclasstype == 1 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[srss] <- "srss"

# regular first 2 years, small, regular
rrsr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 2 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rrsr] <- "rrsr"

# regular first year, small, small, regular
rssr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 1 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rssr] <- "rssr"

# regular first year, small, regular, small
rsrs <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 1

df_students_noaide_4years$class_group[rsrs] <- "rsrs"

# regular first year, small, regular, regular
rsrr <-  df_students_noaide_4years$gkclasstype == 2 & 
  df_students_noaide_4years$g1classtype == 1 & 
  df_students_noaide_4years$g2classtype == 2 &
  df_students_noaide_4years$g3classtype == 2

df_students_noaide_4years$class_group[rsrr] <- "rsrr"
```


```{r}
table(df_students_noaide_4years$class_group)
```
The subset of all small and all regular has good sample size. The subsets of regular to small (rrrs, rrss, rsss) are also sufficient in which we could compare if multiple years leads to higher scores and we could comapre 3rd grade. However subsetting For the other way around small to regular (srrr, ssrr, sssr), the sample is lower. 


## Class Type "Trend" Groups
```{r}
# initalize trend column
df_students_noaide_4years$trend_all <- "Other"

# move column
df_students_noaide_4years <- df_students_noaide_4years %>% 
  relocate(trend_all, .before = class_group) 
```

```{r}
# Change class trend groups for each student

# All 4 years of small vs regular has enough sample size so we can keep those the same
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group == "ssss"] <- "constant_small"
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group == "rrrr"] <- "constant_reg"

# Small to regular trend (srrr, ssrr, sssr)
subset <- c('srrr', 'ssrr', 'sssr')
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group %in% subset] <- "small_to_reg"

# Regular to small trend (rsss, rrss, rrrs)
subset <- c('rsss', 'rrss', 'rrrs')
df_students_noaide_4years$trend_all[df_students_noaide_4years$class_group %in% subset] <- "reg_to_small"
```


## Factor Labels
```{r}
# Set factors and labels

# Urban and free lunch variables for question 1 
df_students_noaide_4years$gksurban <- factor(df_students_noaide_4years$gksurban, labels = c("Inner City", "Suburban", "Rural", "Urban")) # urban


# Gender, race, and free lunch variables for question 2
df_students_noaide_4years$gender <- factor(df_students_noaide_4years$gender, labels = c("Male", "Female")) # gender

df_students_noaide_4years <- subset(df_students_noaide_4years, race %in% c(1,2)) # subset for white and black only
df_students_noaide_4years$race <- factor(df_students_noaide_4years$race, labels = c("White", "Black")) # race

df_students_noaide_4years$g3freelunch <- factor(df_students_noaide_4years$g3freelunch, labels = c("Free", "Non-Free")) # free lunch for grade 3
df_students_noaide_4years$gkfreelunch <- factor(df_students_noaide_4years$gkfreelunch, labels = c("Free", "Non-Free")) # free lunch for kindergarten
```

```{r}
nrow(df_students_noaide_4years)
```


# Early Year Class Size vs Socioeconomic (Question 1)

## Three Way Anova Model 
```{r}
# Urbanicity + free lunch
model_fixed1 <- aov(gktmathss ~ gkclasstype + gkfreelunch + gksurban, data = df_students_noaide_4years)
model_fixed2 <- aov(gktmathss ~ gkclasstype * gkfreelunch + gksurban, data = df_students_noaide_4years)
model_fixed3 <- aov(gktmathss ~ gkclasstype * gksurban + gkfreelunch, data = df_students_noaide_4years)
model_fixed4 <- aov(gktmathss ~ gkclasstype * gksurban * gkfreelunch, data = df_students_noaide_4years)
```

```{r}
summary(model_fixed1)
```

```{r}
summary(model_fixed2)
```

```{r}
summary(model_fixed3)
```

```{r}
summary(model_fixed4)
```

Model equation
$$
Y_{ijkl} = \mu + \alpha_i + \beta_j + \gamma_k + \epsilon_{ijkl} \\

\mu\text{: overall mean} \\
\alpha_i\text{: main effect of class size} \\
\beta_j\text{: main effect of free lunch} \\
\gamma_k\text{: main effect of urbanicity} \\
\epsilon_{ijkl} \sim N(0, \sigma^2)
$$

```{r}
AIC(model_fixed1, model_fixed2, model_fixed3, model_fixed4)
BIC(model_fixed1, model_fixed2, model_fixed3, model_fixed4)
```

```{r}
anova(model_fixed1, model_fixed2, model_fixed3, model_fixed4) 
```

```{r}
# Table df
f_table <- data.frame(
  'Model' = c("Additive", "Interaction (classsize*freelunch)", "Interaction (classsize*urbanicity)", "Interaction (classsize*freelunch*urbanicity)"),
  'Parameters' = c("8", "9", "11", "18"),
  'Residual df' = c("1336", "1335", "1333", "1326"), 
  'AIC' = c("13990", "13992", "13996", "14004"),
  'BIC' = c("14026", "14034", "14048", "14093"),
  'RSS' = c("2619558", "2619513", "2618711", "2608328"),
  'Sum Sq.' = c("", "44.9", "802.4", "10382.6"),
  'F-stat.' = c("", "0.0228", "0.2040", "0.7540"),
  'p-val.' = c("", "0.8799", "0.8155", "0.6261"),
  check.names = FALSE
)

# Create table
kable(f_table, 
      caption = "Table 2: Comparison of Three-way ANOVA Candidate Models by F Test", 
      align = c('l', 'l', 'l', 'l')) %>%
kable_classic(full_width = T, html_font = "Cambria") 
```


F test equation
$$
F = \frac{(RSS_1 - RSS_2)/(df_1 - df_2)}{RSS_2/df_2}
$$


```{r}
anova(model_fixed1)
confint(model_fixed1)
```



# Looking at Switch (Question 2)

## Small class pattern
```{r}
# if we are examining grade 3 math, we have to look at students of similar groups
# Subset for groups of interest: Compliers (constant class size) and Noncompliers (switch)
subset <- c('rsss', 'rrss', 'rrrs', 'rrrr', 'ssss', 'srrr', 'ssrr', 'sssr')
df <- df_students_noaide_4years[df_students_noaide_4years$class_group %in% subset, ] 

nrow(df)
table(df$trend_all)
```

```{r}

# Table df
trend_table <- data.frame(
  'Trend' = c("Constant Regular Class Size", "Constant Small Class Size", "Regular to Small Class Size", "Small to Regular Class Size", "Total"),
  'Sample Size' = c("365", "853", "151", "38", "1407"),
  check.names = FALSE
)

# Create table
kable(trend_table, 
      caption = "Table 1: Trend Variable Sample Size post Filter", 
      align = c('l', 'l', 'l', 'l')) %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  row_spec(5, bold = T, extra_css = "border-top: 1px solid black;")
```


## Model Selection
```{r}
library(lme4)
library(lmerTest)


# Baselines are constant regular class size, white, male, non-free lunch and teacher experience included


# Mixed effects model where school is random intercept examining trend variable
model_mixed1 <- lmer(g3tmathss ~ trend_all + gender + race + g3freelunch + g3tyears + (1 | g3schid), data = df) # additive
model_mixed2 <- lmer(g3tmathss ~ trend_all * g3freelunch + gender + race + g3tyears + (1 | g3schid), data = df) # interaction trend * lunch
model_mixed3 <- lmer(g3tmathss ~ trend_all * race + gender + g3freelunch + g3tyears + (1 | g3schid), data = df) # interaction trend * race
```

Model equation
$$
Y_{ij} = \beta_0 + \beta_1\text{trend} + \beta_2\text{gender} + \beta_3\text{race} + \beta_4\text{freelunch} + \beta_5\text{tyears} + b_{0j}+ \epsilon_{ij}

\\
b_{0j} \sim N(0, \sigma^2_b)\\
\\
\epsilon_{ij} \sim N(0, \sigma^2) 
$$



```{r}
# Looking at larger dataset so use BIC for parsomonious model
BIC(model_mixed1, model_mixed2, model_mixed3)
```


```{r}
# Likelihood ratio test
anova(model_mixed1, model_mixed2)
anova(model_mixed1, model_mixed3)
```


```{r}
# Table df
lrt_table <- data.frame(
  'Model' = c("Additive", "Interaction (trend*freelunch)", "Interaction (trend*race)"),
  'Parameters' = c("10", "13", "13"),
  'AIC' = c("12928", "12927", "12980"),
  'BIC' = c("12980", "12994", "12999"),
  '-2log(L)' = c("12908", "12901", "12906"), 
  'Chisq' = c("", "7.7188", "2.1038"),
  'p-value' = c("", "0.0522", "0.5511"),
  check.names = FALSE
)

# Create table
kable(lrt_table, 
      caption = "Table 3: Comparison of Mixed Effects Candidate Models by Likelihood Ratio Test", 
      align = c('l', 'l', 'l', 'l')) %>%
kable_classic(full_width = F, html_font = "Cambria") 
```


Although the interactions model with lunch had a slightly lower BIC, the likelihood ratio test was not significant indicating the additive model is sufficient.
```{r}
summary(model_mixed1)
```

```{r}
# Table df
modelresults_table <- data.frame(
  'Variable' = c("Intercept", "trendConstantSmall", "trendRegtoSmall", "trendSmalltoReg", "genderFemale", "raceBlack", "Non-free Lunch", "Teacher Years Experience"),
  'Estimate' = c("615.9284", "5.6360", "3.5772", "13.8712", "-1.1203", "-13.4432", "16.5513", "0.2121"),
  'Standard Error' = c("3.9218", "2.3992", "3.7306", "7.2142", "1.9861", "3.4631", "2.4110", "0.1334"),
  'df' = c("485.1381", "1263.1583", "1280.9287", "1204.8663", "1232.0915", "383.7018", "1279.9178", "1171.7302"),
  't-statistic' = c("157.051", "2.349", "0.959", "1.923", "-0.564", "-3.882", "6.865", "1.589"), 
  'p-value' = c("<2e-16", "0.0190", "0.3378", "0.0547", "0.5628", "1.22e-4", "1.04e-11", "0.1123"),
  check.names = FALSE
)

# Create table
kable(modelresults_table, 
      caption = "Table 3: Results of Mixed Effects Additive Model (Fixed Variables)", 
      align = c('l', 'l', 'l', 'l')) %>%
  kable_classic(full_width = F, html_font = "Cambria")


# Table df
modelresults_table2 <- data.frame(
  'Group' = c("School ID (intercept)", "Residual"),
  'Variance' = c("192.2", "1224.7"),
  'Standard Deviation' = c("13.85", "35.00"),
  check.names = FALSE
)

# Create table
kable(modelresults_table2, 
      caption = "Table 4: Results of Mixed Effects Additive Model (Random Variable)", 
      align = c('l', 'l', 'c', 'm')) %>%
  kable_classic(full_width = F, html_font = "Cambria")

```




```{r}
anova(model_mixed1)
confint(model_mixed1)
```




## Noncomplier Random (prob scrap)
```{r}
# Compares baselines in kindergarden between compliers and non-compliers  
t.test(gktmathss ~ trend_all, data = subset(df, trend_all %in% c("constant_reg", "reg_to_small"))) 
t.test(gktmathss ~ trend_all, data = subset(df, trend_all %in% c("constant_small", "small_to_reg"))) 
```




# Sensitivity Analysis

## Intraclass Correlation Coefficient (ICC)
```{r}
# ICC for trend model
var_comp <- as.data.frame(VarCorr(model_mixed1))
v_school <- var_comp$vcov[1]
v_residual <- var_comp$vcov[2]

v_school / (v_school + v_residual)
```

Around 13.56% of student's score is due to school effects when considered the model for trend

## Assumptions
```{r}
# residuals vs fitted plot
plot(model_mixed1) 
```



